\chapter{Literature Review} \label{lit-review}

\section{A History of Deep Learning}
\subsection{Cybernetics}
Deep learning originates from the field of cybernetics \cite{deeplearningbook}, which aimed to create computational models of biological learning in the brain. The first effort to model neuronal activity was by Mculloch and Pitts, who introduced the McCulloch-Pitts neuron (MCP) in 1943 \cite{mculloch-pitts}. The MCP utilized propositional logic to form "nervous nets," computational structures that could be composed to model the sequential firing of neurons over time. In their work, Mculloch and Pitts modeled outputs as binary, either active or inactive based on the activity of the afferent synapses that connected to them. Synaptic connections could be either excitatory or inhibitory. Although MCP provided the first framework for modeling brain activity with a computational model, it was not without limitations. The model did not have a true "learning mechanism." Instead, all parameters were set by hand. 

Building on the work of McCulloch and Pitts, Rosenblatt introduced the perceptron in 1958 \cite{rosenblatt-perceptron}. Unlike the MCP, the perceptron was built on a foundation of probability theory, utilizing the principle of statistical separability to sort inputs into distinct categories. Furthermore, Rosenblatt experimentally analyzed the perceptron, showing that it could distinguish markings on the left and right sides of a punch card from each other. This marked the first definition and implementation of a learning rule for artificial intelligence. Rosenblatt's 


\subsection{Connectionism}

\subsection{Deep Learning}